Engineering infinite-range SU($n$) interactions with spin-orbit coupled fermions in an optical lattice

### 2021-04-28 sun_phases/spin_model_paper/main.tex
last major updates before blue-forming the SU(n) spin model paper

\title{Engineering an infinite-range SU($n$) spin model with ultracold fermions in an optical lattice}
\author{Michael A. Perlin}
% \orcid{https://orcid.org/0000-0002-9316-1596}
\email{mika.perlin@gmail.com}
\author{Diego Barbarena}
\affiliation{\JILA}
\affiliation{\CTQM}
\author{Robert J.~Lewis-Swan}
\affiliation{\UO}
\affiliation{\CQRT}
% \orcid{https://orcid.org/0000-0002-0201-281X}
\author{Ana Maria Rey}
% \orcid{http://orcid.org/0000-0001-7176-9413}
\affiliation{\JILA}
\affiliation{\CTQM}

@@

\begin{abstract}
We study multilevel fermions in an optical lattice, with dynamics described by the Hubbard model with on site SU($n$)-symmetric interactions.
We show that in an appropriate parameter regime this system can be mapped onto a multilevel spin model with all-to-all SU($n$)-symmetric couplings.
Raman pulses that address internal spin states modify the atomic dispersion relation and induce spin-orbit coupling, which can act as a synthetic inhomogeneous magnetic field that competes with the SU($n$) exchange interactions.
We investigate the mean-field dynamical phase diagram of the resulting model as a function of $n$ and different initial configurations that are accessible with Raman pulses.
Consistent with previous studies for $n=2$, we find that for some initial states the spin model exhibits two distinct dynamical phases that obey simple scaling relations with $n$.
Moreover, for $n>2$ we find that dynamical behavior can be highly sensitive to initial intra-spin coherences.
Our predictions are readily testable in current experiments with ultracold alkalike-earth(-like) atoms.
\end{abstract}

@@

SU($n$) symmetries play an important role in physics.
Underpinning much of high energy physics, the SU($n$) gauge theory known as Yang-Mills theory is central to our understanding of the electroweak and strong forces.
Extensions of Yang-Mills and SU($n$) symmetry feature in the most well-studied examples of holographic duality \cite{maldacena1999largen} and the connection between entanglement and gravity \cite{ryu2006holographic} through the anti-de Sitter/conformal field theory (AdS/CFT) correspondence.
In a condensed matter setting, SU($2$) appears ubiquitously as a symmetry of the Hubbard model, with important consequences for the study of quantum magnetism and high temperature superconductivity \cite{lee2006doping}.
The extension of SU($2$) Hubbard and spin models to SU($n$) has led to predictions of exotic phases of matter such as valence bond solids \cite{read1989valencebond, rokhsar1990quadratic, kaul2012lattice, hermele2011topological} and chiral spin liquids \cite{hermele2009mott, hermele2011topological, chen2016syntheticgaugefield, nataf2016chiral}, as well as the potential to perform universal topological quantum computation \cite{freedman2004class, nayak2008nonabelian} and other phenomena \cite{nataf2014exact, nataf2016exact}.
Furthermore, disordered SU($n$) spin models have opened analytically tractable avenues for studying quantum chaos and information scrambling \cite{sachdev1993gapless, bentsen2019integrable}.

The tremendous theoretical significance of SU($n$) symmetries makes it all the more exciting that they appear naturally in experimental atomic, molecular, and optical (AMO) platforms with exquisite degrees of microscopic control.
This symmetry arises through the independence of atomic orbital and interaction parameters on the $n$ nuclear spin states of alkaline-earth(-like) atoms, with e.g.~$n=10$ for ${}^{87}$Sr \cite{wu2003exact, cazalilla2009ultracold, gorshkov2010twoorbital, cazalilla2014ultracold}.
As a result, AMO experiments can directly probe the role of SU($n$) interactions in controllable settings.
Recent progress includes the experimental observations of bosonization and faster thermalization in multi-component Fermi gas with SU($n$) symmetry \cite{hazzard2012hightemperature, messio2012entropy, bonnes2012adiabatic, stellmer2013degenerate, yip2014theory, pagano2014onedimensional, choudhury2020collective, song2020evidence, sonderhouse2020thermodynamics}, SU($n$) Hubbard phases and phase transitions \cite{taie2012su, hofrichter2016direct, taie2020observation}, two-orbital SU($n$) magnetism \cite{cappellini2014direct, scazza2014observation, zhang2014spectroscopic, beverland2016realizing}, and multi-body SU($n$)-symmetric interactions \cite{goban2018emergence, perlin2019effective}.

@@

Finally, we take a moment to discuss individual $n$-level spins.
The state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of a $n$-level spin can be represented by a probability distribution\footnote{Strictly speaking, in order to make $Q_\psi$ (and $Q_{\hat\rho}$) a probability distribution, one has to divide it by a normalization factor $\frac{4\pi}{d}$.
We omit this normalization factor for convenience.} $Q_\psi$ on the sphere $\SS^2$.
The value $Q_\psi\p{\v v}$ at a point $\v v\in\SS^2$ equal to the overlap of $\ket\psi$ with a pure state $\ket{\v v}$ maximally polarized along $\v v$: $Q_\psi\equiv\abs{\bk{\v v|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\hat\rho$, this distribution is defined by $Q_{\hat\rho}\equiv\bk{\v v|\hat\rho|\v v}$\footnote{The dimension $n$ is reflected in $Q_{\hat\rho}$ by the fact that it can be decomposed into spherical harmonics $Y_{\ell m}$ with degree $\ell<n$ (that is, $Q_{\hat\rho}$ is ``low resolution'' or ``band-limited'' at degree $n$) \cite{perlin2020qudit}, as well as in the normalization $\int_{\v v\in\SS^2}\dd\v v\,Q_{\hat\rho}\p{\v v}=\frac{4\pi}{d}$.}.
Closely related spherical representations of multilevel spin states (and operators) are discussed in Ref.~\cite{li2013weylwignermoyal}.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{qudit_rep.pdf}
\caption{
Whereas the state of a two-level spin (qubit) can be represented by a point on (or inside) the Bloch sphere, the state of a $n$-level spin is more generally represented by a probability distribution on the sphere, decomposable into spherical harmonics $Y_{\ell m}$ of degree $\ell<n$.
The distribution shown for $n=10$ corresponds to a Haar-random pure state.
}
\label{fig:spin_dist}
\end{figure}

In fact, $Q_{\hat\rho}$ is essentially the Husimi-$Q$ function commonly used (e.g.~in the spin squeezing community \cite{ma2011quantum}) to represent collective states of spin-$\frac{1}{2}$ particles in the Dicke manifold.
This correspondence is made precise by identifying the Hilbert space of a single $n$-level spin with the Dicke manifold of $n-1$ spin-$\frac{1}{2}$ particles.
The Hilbert space of a 10-level (spin-$\frac92$) nucleus of a ${}^{87}$Sr atom, for example, can be thought of as the Dicke manifold of 9 spin-$\frac{1}{2}$ particles (``excess'' protons and neutrons).
We will freely use this correspondence to import standard results and constructions for the Dicke manifold (e.g.~the spin operators $\s_\x,\s_\y,\s_\z$) into the context of a multilevel spin.

@@

We now consider a specific three-laser driving scheme with a geometry sketched in Figure \ref{fig:3LD}.
Here the lattice lies in the y-z plane at an angle $\theta$ to the z axis, oriented along $\v\ell=(0,\sin\theta,\cos\theta)$; two counter-propagating right-circularly polarized lasers point in opposite directions along the z axis, $\v v_\pm=(0,0,\pm1)$; and a third linearly-polarized laser points along the x axis, $\v v_0=(1,0,0)$.
Choosing $\v v_+$ (i.e.~the z axis) as the spin quantization axis, we can define new drive amplitudes $\Omega_0\equiv-(\Omega_{\v v_0,+}+\Omega_{\v v_0,-})$ and $\Omega_\pm\equiv\pm\Omega_{\v v_\pm,+}$, as well as a SOC angle $\phi\equiv\kappa\v v_+\cdot\v\ell=\kappa\cos\theta$, in terms of which the drive Hamiltonian becomes
\begin{align}
  \H_{\t{3LD}}^{\t{full}}
  = \sum_{j,m} \Omega_m
  \p{e^{-\i m\phi j} \s_{mj} \otimes\op{\e}{\g}_j + \t{h.c.}}
  + \Delta \hat N_\e,
\end{align}
where $m\in\set{+1,0,-1}$ indexes the laser pointing along $\v v_m$; and $\s_{0,j}\equiv \s_{\z,j}$ for shorthand.
In the far-detuned limit $\abs{\Delta}\gg\abs{\Omega_m}$, a second-order perturbative treatment of electronic excitations ($\ket\e$) yields an effective drive Hamiltonian that only addresses ground-state nuclear spins.
After additionally making the gauge transformation $\s_{mj}\to e^{\i m\phi j}\s_{mj}$ (equivalently $\c_{j\mu}^\dag\to e^{\i\phi\mu j} \c_{j\mu}^\dag$), the drive Hamiltonian then becomes
\begin{align}
  \H_{\t{3LD}} = \sum_j \H_{\t{3LD},j}^{\t{single}},
  \label{eq:H_3LD}
\end{align}
where $\H_{\t{3LD},j}^{\t{single}}$ denotes the action of $\H_{\t{3LD}}^{\t{single}}$ on spin $j$:
\begin{multline}
  \H_{\t{3LD}}^{\t{single}}
  = \tilde\Omega_+ \tilde\Omega_- \s_\z
  + \tilde\Omega_0 \tilde\Omega_- \s_\x
  + \tilde\Omega_0 \tilde\Omega_+ (\s_\z \s_\x  + \s_\x \s_\z) \\
  - \tilde\Omega_0^2 \s_\z^2 - \tilde\Omega_+^2 \s_\x^2
  - \tilde\Omega_-^2 \s_\y^2,
  \label{eq:H_3LD_single}
\end{multline}
with
\begin{align}
  \tilde\Omega_0 \equiv -\f{\Omega_0}{\sqrt\Delta},
  &&
  \tilde\Omega_\pm \equiv \f{\Omega_+\pm\Omega_-}{\sqrt\Delta},
\end{align}
where we have made the simplifying assumption that all drive amplitudes are real to arrive at the form of $\H_{\t{3LD}}^{\t{single}}$ in Eq.~\eqref{eq:H_3LD_single}.
We relax the assumption of real drive amplitudes in Appendix \ref{sec:full_drive}.

@@

Finally, we comment on the preparation of initial states.
Initial states are nominally prepared in the ``lab frame'', and must be transformed according to the gauge transformation $\c_{j\mu}^\dag\to e^{\i\phi\mu j} \c_{j\mu}^\dag$ prior to evolution under the three-laser drive $\H_{\t{3LD}}$ in Eq.~\eqref{eq:H_3LD}, which is expressed in the ``gauge frame''.
We assume the capability to prepare an initial state in which all spins are maximally polarized along the z axis, i.e.~$\ket{\z}^{\otimes N}=\ket{s}^{\otimes N}$, which is unaffected by the gauge transformation (up to a global phase).
The three-laser then allows us to rotate this state into one that is polarized along any spatial axis (in the gauge frame).
In addition, when $n>2$ the three-laser drive allows us to prepare product states with nontrivial intra-spin correlations.
For example, when $n$ is even we can prepare an $N$-fold product of the ``kitten'' state
\begin{align}
  e^{-\i\frac{\pi}{2}\p{\s_\y + \s_\y^2}} \ket{s}
  \propto \ket{s} + \ket{-s}.
\end{align}
This state has a vanishing mean spin vector, $\bk{\s_\x}=\bk{\s_\y}=\bk{\s_\z}=0$, but variances $\bk{\s_\x^2}=\bk{\s_\y^2}=s/2$ and $\bk{\s_\z^2}=s^2$.

@@

The two-level SOC drive in Eq.~\eqref{eq:drive_2} has been implemented with an external laser that couples the two electronic states of nuclear-spin-polarized atoms, with $\dn$ ($\up$) indexing the ground (excited) electronic state \cite{wall2016synthetic, livi2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
In contrast, the drive we considered in Section \ref{sec:controls} addresses electronic excitations off-resonantly, inducing an effective Hamiltonian in the ground-state hyperfine manifold with spin projections $\mu\in\set{s,s-1,\cdots,-s}$ (a similar one-laser scheme was used to study SOC using a subspace of the ground-state manifold in Ref.~\cite{mancini2015observation}).
Nonetheless, both the two-level drive in Eq.~\eqref{eq:drive_2} and the $n$-level drive in Eq.~\eqref{eq:H_3LD} become homogeneous (i.e.~independent of the spatial mode index $j$ or $q$) and independent of the SOC angle $\phi$ after the same spin-symmetric gauge transformation\footnote{Note that the ``asymmetric'' gauge transformation $(\c_{j,\up}^\dag,\c_{j,\dn}^\dag)\to(e^{\i\phi j} \c_{j,\up}^\dag,\c_{j,\dn}^\dag)$, sometimes performed in the two-state SOC literature, does not generalize as nicely to $n>2$.} $\c_{j\mu}^\dag\to e^{\i\phi\mu j} \c_{j\mu}^\dag$.

Of course, spin-orbit coupling cannot be ``gauged away'' entirely.
Making a gauge transformation to simplify the drive comes at the cost of making the kinetic energy in Eq.~\eqref{H_kin_start} spin-dependent, taking
\begin{align}
  \H_{\t{kin}} \to \H_{\t{kin}}^{(\phi)}
  \equiv -2J \sum_q \cos\p{q+\mu\phi} \s_{\mu\mu q},
\end{align}
as visualized in Figure \ref{fig:soc_panels}.
To better interpret this Hamiltonian, we can write it in the form
\begin{align}
  \H_{\t{kin}}^{(\phi)}
  = -2J \sum_q
  \sp{\cos\p{q} \hat w_{+,q}^{(\phi)} - \sin\p{q} \hat w_{+,q}^{(\phi)}},
\end{align}
where
\begin{align}
  \hat w_{+,q}^{(\phi)} &\equiv \sum_\mu \cos\p{\mu\phi} \s_{\mu\mu q}, \\
  \hat w_{-,q}^{(\phi)} &\equiv \sum_\mu \sin\p{\mu\phi} \s_{\mu\mu q}.
\end{align}
For two-level spins with $\mu=\pm\frac12$, $\hat w_{+,q}^{(\phi)}$ is proportional to the identity operator and $\hat w_{+,q}^{(\phi)} = 2\sin\p{\phi/2} \s_{\z,q}$, so the kinetic Hamiltonian in the gauge frame describes a (synthetic) inhomogeneous magnetic field:
\begin{align}
  \left. \H_{\t{kin}}^{(\phi)} \right|_{n=2}
  = 4 J \sin\p{\phi/2} \sum_q \sin\p{q} \s_{\z,q}.
\end{align}
When $n>2$, an inhomogeneous magnetic field is likewise recovered in the weak SOC limit $s\phi\ll1$, in which case
\begin{align}
  \left. \H_{\t{kin}}^{(\phi)} \right|_{s\phi\ll1}
  = 2J\phi \sum_q \sin\p{q} \s_{\z,q} + O\p{(s\phi)^2}.
\end{align}
For larger $\phi$, this Hamiltonian acquires terms with higher powers of $\s_{\z,q}$, up to $\s_{\z,q}^{n-1}$.

\begin{figure}
\centering
\includegraphics{soc_panels.pdf}
\caption{
Spin-orbit coupling for 2-level ({\bf a},{\bf c}) and 4-level ({\bf b},{\bf d}) spins.
Colors indicate different spin projections $\mu$.
In the ``lab frame'' ({\bf a},{\bf b}), kinetic energy is insensitive to spin, but the drive imparts a momentum kick $q\to q+\phi$ into any atom that absorbs a photon.
Changing to the ``gauge frame'' ({\bf c},{\bf d}), essentially by shifting the momentum label $q$ for each spin state $\mu$, makes the drive diagonal in the momentum index, but comes at the cost of making kinetic energy spin-dependent.
}
\label{fig:soc_panels}
\end{figure}

@@

We now study the dynamical behavior of the SOC spin Hamiltonian $H_\spin$ in Eq.~\eqref{eq:H_spin}, and henceforth work exclusively in the ``gauge frame'' of $\H_\spin$ and the three-laser drive $\H_{\t{3LD}}$ in Eq.~\eqref{eq:H_3LD}.
The general idea will be a Ramsey-type setup in which we first prepare an initial state with the three-laser drive (using fast pule sequences), let this state evolve freely for some time under $\H_\spin$, and then again use the three-laser drive to map observables of interest onto spin projection measurements (e.g.~with spin qudit tomography \cite{newton1968measurability, perlin2020qudit}).
At the mean-field (MF) level, up to constant energy shifts the undriven spin Hamiltonian becomes
\begin{align}
  \H_\MF = u \sum_q\sp{-\bk{\bar{\v s}}\cdot\v\s_q
    + h \sin\p{q} \s_{\z,q}}.
  \label{eq:H_MF}
\end{align}
where $\bar{\v s}\equiv\frac1N\sum_q\v\s_q$ is the average spin matrix, and $h\equiv 2J\phi/u$ is a dimensionless strength of the inhomogeneous magnetic field.
We assume that all momenta $q\in\ZZ_N\times 2\pi/N$ are occupied.
Fixing the atom number $N$, the spin Hamiltonian has one free parameter, $h$, which determines the relative strength of the single-particle and interaction terms.
One should therefore expect distinct dynamical behaviors when $h\ll1$, in which case strong spin-locking interactions should give rise to a long-range ordered phase, as opposed to $h\gg1$, in which case long-range order should be destroyed by the strong inhomogeneous magnetic field \cite{smale2019observation}.

@@

Our numerical simulations of mean-field dynamics are performed with a Schwinger boson decomposition of spin operators: $\s_{\mu\nu q} = \b_{\mu q}^\dag\b_{\nu q}$.
This decomposition requires no approximations, and reduces the number of variables to keep track of by a factor of $\sim n$.
See Appendices \ref{sec:MFT} and \ref{sec:bosons} additional details about our numerical simulations and the Schwinger boson equations of motion.

>>>>

\title{Engineering infinite-range SU($n$) interactions with spin-orbit coupled fermions in an optical lattice}
\author{Michael A. Perlin}
% \orcid{https://orcid.org/0000-0002-9316-1596}
\email{mika.perlin@gmail.com}
\author{Diego Barberena}
\affiliation{\JILA}
\affiliation{\CTQM}
\author{Robert J.~Lewis-Swan}
\affiliation{\UO}
\affiliation{\CQRT}
% \orcid{https://orcid.org/0000-0002-0201-281X}
\author{Ana Maria Rey}
% \orcid{http://orcid.org/0000-0001-7176-9413}
\affiliation{\JILA}
\affiliation{\CTQM}

@@

\begin{abstract}
We study multilevel fermions in an optical lattice described by the Hubbard model with on site SU($n$)-symmetric interactions.
We show that in an appropriate parameter regime this system can be mapped onto a spin model with all-to-all SU($n$)-symmetric couplings.
Raman pulses that address internal spin states modify the atomic dispersion relation and induce spin-orbit coupling, which can act as a synthetic inhomogeneous magnetic field that competes with the SU($n$) exchange interactions.
We investigate the mean-field dynamical phase diagram of the resulting model as a function of $n$ and different initial configurations that are accessible with Raman pulses.
Consistent with previous studies for $n=2$, we find that for some initial states the spin model exhibits two distinct dynamical phases that obey simple scaling relations with $n$.
Moreover, for $n>2$ we find that dynamical behavior can be highly sensitive to initial intra-spin coherences.
Our predictions are readily testable in current experiments with ultracold alkalike-earth(-like) atoms.
\end{abstract}

@@

SU($n$) symmetries play an important role in physics.
Underpinning much of high energy physics, the SU($n$) gauge theory known as Yang-Mills theory is central to our understanding of the electroweak and strong forces.
Extensions of Yang-Mills and SU($n$) symmetry feature in the most well-studied examples of holographic duality \cite{maldacena1999largen} and the connection between entanglement and gravity \cite{ryu2006holographic} through the anti-de Sitter/conformal field theory (AdS/CFT) correspondence.
In a condensed matter setting, SU($2$) appears ubiquitously as a symmetry of the Hubbard model, with important consequences for the study of quantum magnetism and high temperature superconductivity \cite{lee2006doping}.
The extension of SU($2$) Hubbard and spin models to SU($n$) has led to predictions of exotic phases of matter such as valence bond solids \cite{read1989valencebond, rokhsar1990quadratic, kaul2012lattice, hermele2011topological} and chiral spin liquids \cite{hermele2009mott, hermele2011topological, chen2016syntheticgaugefield, nataf2016chiral}, as well as the potential to perform universal topological quantum computation \cite{freedman2004class, nayak2008nonabelian} and other phenomena \cite{nataf2014exact, nataf2016exact}.
Furthermore, disordered SU($n$) spin models have opened analytically tractable avenues for studying quantum chaos and information scrambling \cite{sachdev1993gapless}.

The tremendous theoretical significance of SU($n$) symmetries makes it all the more exciting that they appear naturally in experimental atomic, molecular, and optical (AMO) platforms with exquisite degrees of microscopic control.
This symmetry arises through the independence of atomic orbital and interaction parameters on the $n$ nuclear spin states of alkaline-earth(-like) atoms, with e.g.~$n=10$ for ${}^{87}$Sr \cite{wu2003exact, cazalilla2009ultracold, gorshkov2010twoorbital, cazalilla2014ultracold}.
As a result, AMO experiments can directly probe the role of SU($n$) interactions in controllable settings.
Recent progress includes studies of the thermodynamic properties of SU($n$) fermionic gases \cite{hazzard2012hightemperature, bonnes2012adiabatic, stellmer2013degenerate, yip2014theory, pagano2014onedimensional, choudhury2020collective, song2020evidence, sonderhouse2020thermodynamics}, SU($n$) Hubbard phases and phase transitions \cite{taie2012su, hofrichter2016direct, taie2020observation}, single- \cite{messio2012entropy} and two-orbital \cite{cappellini2014direct, scazza2014observation, zhang2014spectroscopic, beverland2016realizing} SU($n$) magnetism, and multi-body SU($n$)-symmetric interactions \cite{goban2018emergence, perlin2019effective}.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{qudit_rep.pdf}
\caption{
Whereas the state of a two-level spin (qubit) can be represented by a point on (or inside) the Bloch sphere, the state of a $n$-level spin is more generally represented by a probability distribution on the sphere.
The distribution shown for $n=10$ corresponds to a Haar-random pure state.
}
\label{fig:spin_dist}
\end{figure}

Finally, we take a moment to discuss individual $n$-level spins.
The state of a two-level spin, or a qubit, is commonly represented by a point on (or within) the Bloch sphere.
More generally, the state $\ket\psi$ of a $n$-level spin can be represented by a quasi-probability distribution $Q_\psi$ on the sphere (commonly known as the Husimi-$Q$ function, e.g.~in the spin-squeezing community \cite{ma2011quantum}).
The value $Q_\psi\p{\v v}$ at a point $\v v$ on the sphere is equal to the overlap of $\ket\psi$ with a pure state $\ket{\v v}$ maximally polarized in the direction of $\v v$: $Q_\psi\equiv\abs{\bk{\v v|\psi}}^2$ (see Figure \ref{fig:spin_dist}).
In the case of a mixed state $\hat\rho$, this distribution is defined by $Q_{\hat\rho}\equiv\bk{\v v|\hat\rho|\v v}$.
Closely related spherical representations of multilevel spin states and operators are discussed in Refs.~\cite{dowling1994wigner, li2013weylwignermoyal}.
In practice, it is conceptually useful to identify the Hilbert space of a single $n$-level spin with the Dicke manifold of $n-1$ spin-$\frac12$ particles.

@@

We now consider a specific three-laser driving scheme with a geometry sketched in Figure \ref{fig:3LD}.
Here the lattice lies in the y-z plane at an angle $\theta$ to the z axis, oriented along $\v\ell=(0,\sin\theta,\cos\theta)$; two counter-propagating right-circularly polarized lasers point in opposite directions along the z axis, $\v v_\pm=(0,0,\pm1)$; and a third linearly-polarized laser points along the x axis, $\v v_0=(1,0,0)$.
Choosing $\v v_+$ (i.e.~the z axis) as the spin quantization axis, we can define new drive amplitudes $\Omega_0\equiv-(\Omega_{\v v_0,+}+\Omega_{\v v_0,-})$ and $\Omega_\pm\equiv\pm\Omega_{\v v_\pm,+}$, as well as a SOC angle $\phi\equiv\kappa\v v_+\cdot\v\ell=\kappa\cos\theta$, in terms of which the drive Hamiltonian becomes
\begin{align}
  \H_{\t{3LD}}^{\t{full}}
  = \sum_{j,m} \Omega_m
  \p{e^{-\i m\phi j} \s_{mj} \otimes\op{\e}{\g}_j + \t{h.c.}}
  + \Delta \hat N_\e,
\end{align}
where $m\in\set{+1,0,-1}$ indexes the laser pointing along $\v v_m$; $\s_{0,j}\equiv\s_{\z,j}$ for shorthand; and $\s_{\z,j},\s_{+,j}\s_{-,j}$ are standard axial, spin-raising, and spin-lowering operators for spin $j$.
In the far-detuned limit $\abs{\Delta}\gg\abs{\Omega_m}$, a second-order perturbative treatment of electronic excitations ($\ket\e$) yields an effective drive Hamiltonian that only addresses ground-state nuclear spins.
After additionally making the gauge transformation $\s_{mj}\to e^{\i m\phi j}\s_{mj}$ (equivalently $\c_{j\mu}^\dag\to e^{\i\phi\mu j} \c_{j\mu}^\dag$), the drive Hamiltonian then becomes
\begin{align}
  \H_{\t{3LD}} = \sum_j \H_{\t{3LD},j}^{\t{single}},
  \label{eq:H_3LD}
\end{align}
where $\H_{\t{3LD},j}^{\t{single}}$ denotes the action of $\H_{\t{3LD}}^{\t{single}}$ on spin $j$:
\begin{multline}
  \H_{\t{3LD}}^{\t{single}}
  = \tilde\Omega_+ \tilde\Omega_- \s_\z
  + \tilde\Omega_0 \tilde\Omega_- \s_\x
  + \tilde\Omega_0 \tilde\Omega_+ (\s_\z \s_\x  + \s_\x \s_\z) \\
  - \tilde\Omega_0^2 \s_\z^2 - \tilde\Omega_+^2 \s_\x^2
  - \tilde\Omega_-^2 \s_\y^2,
  \label{eq:H_3LD_single}
\end{multline}
with
\begin{align}
  \tilde\Omega_0 \equiv -\f{\Omega_0}{\sqrt\Delta},
  &&
  \tilde\Omega_\pm \equiv \f{\Omega_+\pm\Omega_-}{\sqrt\Delta},
\end{align}
where we have made the simplifying assumption that all drive amplitudes are real to arrive at the form of $\H_{\t{3LD}}^{\t{single}}$ in Eq.~\eqref{eq:H_3LD_single}.
We relax the assumption of real drive amplitudes in Appendix \ref{sec:full_drive}.

@@

Finally, we comment on the preparation of initial states.
Initial states are nominally prepared in the ``lab frame'', and must be transformed according to the gauge transformation $\c_{j\mu}^\dag\to e^{\i\phi\mu j} \c_{j\mu}^\dag$ prior to evolution under the three-laser drive $\H_{\t{3LD}}$ in Eq.~\eqref{eq:H_3LD}, which is expressed in the ``gauge frame''.
We assume the capability to prepare an initial state in which all spins are maximally polarized along the z axis, i.e.~$\ket{\z}^{\otimes N}=\ket{s}^{\otimes N}$, which is unaffected by the gauge transformation (up to a global phase).
The three-laser then allows us to rotate this state into one that is polarized along any spatial axis (in the gauge frame).
In addition, when $n>2$ the three-laser drive allows us to prepare product states with nontrivial intra-spin correlations.
For example, when $n$ is even we can prepare an $N$-fold product of the ``kitten'' state
\begin{align}
  e^{-\i\frac{\pi}{2}\p{\s_\y + \s_\y^2}} \ket{s}
  \stackrel{n\,\t{even}}{\propto} \ket{s} + \ket{-s}.
\end{align}
This state has a vanishing mean spin vector, $\bk{\s_\x}=\bk{\s_\y}=\bk{\s_\z}=0$, but variances $\bk{\s_\x^2}=\bk{\s_\y^2}=s/2$ and $\bk{\s_\z^2}=s^2$.

@@

The two-level SOC drive in Eq.~\eqref{eq:drive_2} has been implemented with an external laser that couples the two electronic states of nuclear-spin-polarized atoms, with $\dn$ ($\up$) indexing the ground (excited) electronic state \cite{wall2016synthetic, livi2016synthetic, kolkowitz2016spinorbitcoupled, bromley2018dynamics, he2019engineering}.
In contrast, the drive we considered in Section \ref{sec:controls} addresses electronic excitations off-resonantly, inducing an effective Hamiltonian in the ground-state hyperfine manifold with spin projections $\mu\in\set{s,s-1,\cdots,-s}$ (a similar scheme was used to study SOC in a subspace of the ground-state manifold in Ref.~\cite{mancini2015observation}).
Nonetheless, both the two-level drive in Eq.~\eqref{eq:drive_2} and the $n$-level drive in Eq.~\eqref{eq:H_3LD} become homogeneous (i.e.~independent of the spatial mode index $j$ or $q$) and independent of the SOC angle $\phi$ after the same spin-symmetric gauge transformation\footnote{Note that the ``asymmetric'' gauge transformation $(\c_{j,\up}^\dag,\c_{j,\dn}^\dag)\to(e^{\i\phi j} \c_{j,\up}^\dag,\c_{j,\dn}^\dag)$, sometimes performed in the two-state SOC literature, does not generalize as nicely to $n>2$.} $\c_{j\mu}^\dag\to e^{\i\phi\mu j} \c_{j\mu}^\dag$.

Of course, spin-orbit coupling cannot be ``gauged away'' entirely.
Making a gauge transformation to simplify the drive comes at the cost of making the kinetic energy in Eq.~\eqref{H_kin_start} spin-dependent, taking
\begin{align}
  \H_{\t{kin}} \to \H_{\t{kin}}^{(\phi)}
  \equiv -2J \sum_q \cos\p{q+\mu\phi} \s_{\mu\mu q},
\end{align}
as visualized in Figure \ref{fig:soc_panels}.
To better interpret this Hamiltonian, we can write it in the form
\begin{align}
  \H_{\t{kin}}^{(\phi)}
  = -2J \sum_q
  \sp{\cos\p{q} \hat w_{+,q}^{(\phi)} - \sin\p{q} \hat w_{-,q}^{(\phi)}},
\end{align}
where
\begin{align}
  \hat w_{+,q}^{(\phi)} &\equiv \sum_\mu \cos\p{\mu\phi} \s_{\mu\mu q}, \\
  \hat w_{-,q}^{(\phi)} &\equiv \sum_\mu \sin\p{\mu\phi} \s_{\mu\mu q}.
\end{align}
For two-level spins with $\mu=\pm\frac12$, $\hat w_{+,q}^{(\phi)}$ is proportional to the identity operator and $\hat w_{+,q}^{(\phi)} = 2\sin\p{\phi/2} \s_{\z,q}$, so the kinetic Hamiltonian in the gauge frame describes a (synthetic) inhomogeneous magnetic field:
\begin{align}
  \left. \H_{\t{kin}}^{(\phi)} \right|_{n=2}
  = 4 J \sin\p{\phi/2} \sum_q \sin\p{q} \s_{\z,q}.
\end{align}
When $n>2$, an inhomogeneous magnetic field is likewise recovered in the weak SOC limit $s\phi\ll1$, in which case
\begin{align}
  \left. \H_{\t{kin}}^{(\phi)} \right|_{s\phi\ll1}
  = 2J\phi \sum_q \sin\p{q} \s_{\z,q} + O\p{(s\phi)^2}.
\end{align}
For larger $\phi$, this Hamiltonian acquires terms with higher powers of $\s_{\z,q}$, up to $\s_{\z,q}^{n-1}$.

\begin{figure}
\centering
\includegraphics{soc_panels.pdf}
\caption{
Spin-orbit coupling for 2-level ({\bf a},{\bf c}) and 4-level ({\bf b},{\bf d}) spins.
Colors indicate different spin projections $\mu$.
In the ``lab frame'' ({\bf a},{\bf b}), kinetic energy is insensitive to spin, but a spin transition $\mu\to\mu+1$ is accompanied by a momentum kick $q\to q+\phi$ from the drive.
Changing into the ``gauge frame'' ({\bf c},{\bf d}), essentially by shifting the momentum label $q$ for each spin state $\mu$, makes the drive diagonal in the momentum index, but comes at the cost of making kinetic energy spin-dependent.
}
\label{fig:soc_panels}
\end{figure}

@@

We now study the dynamical behavior of the SOC spin Hamiltonian $H_\spin$ in Eq.~\eqref{eq:H_spin}, and henceforth work exclusively in the ``gauge frame'' of $\H_\spin$ and the three-laser drive $\H_{\t{3LD}}$ in Eq.~\eqref{eq:H_3LD}.
We use a Ramsey-like setup wherein we prepare an initial state with the three-laser drive (using fast pule sequences), then let the state evolve freely for some time under $\H_\spin$, and finally apply again the three-laser drive to map observables of interest onto spin projection measurements (e.g.~with spin qudit tomography \cite{newton1968measurability, perlin2020qudit}).
At the mean-field (MF) level, the undriven spin Hamiltonian (neglecting constant energy shifts) becomes
\begin{align}
  \H_\MF = u \sum_q\sp{-\bk{\bar{\v s}}\cdot\v\s_q
    + h \sin\p{q} \s_{\z,q}}.
  \label{eq:H_MF}
\end{align}
where $\bar{\v s}\equiv\frac1N\sum_q\v\s_q$ is the average spin matrix, and $h\equiv 2J\phi/u$ is a dimensionless strength of the inhomogeneous magnetic field.
We assume that all momenta $q\in\ZZ_N\times 2\pi/N$ are occupied.
Fixing the atom number $N$, the spin Hamiltonian has one free parameter, $h$, which determines the relative strength of the single-particle and interaction terms.
One should therefore expect distinct dynamical behaviors when $h\ll1$, in which case strong spin-locking interactions should give rise to a long-range ordered phase, as opposed to $h\gg1$, in which case long-range order should be destroyed by the strong inhomogeneous magnetic field \cite{smale2019observation}.

@@

Our numerical simulations of mean-field dynamics are performed with a Schwinger boson decomposition of spin operators: $\s_{\mu\nu q} = \b_{\mu q}^\dag\b_{\nu q}$.
This decomposition requires no approximations, and reduces the number of variables to keep track of by a factor of $\sim n$.
See Appendices \ref{sec:MFT} and \ref{sec:bosons} for additional details about our numerical simulations and the Schwinger boson equations of motion.
